---
title: "Geography 176A"
author: "[Catherine Rauch](https://catherinerauch.github.io)"
subtitle: 'Lab 05'
output:
  html_document:
    theme: spacelab
---


***

### Libraries

```{r setup, message=FALSE}

library(raster) # Raster Data handling
library(tidyverse) # Data Manipulation
library(getlandsat) # keyless Landsat data (2013-2017)
library(sf) # Vector data processing
library(mapview) # Rapid Interactive visualization
library(kableExtra)
library(sp)

```

*** 
### Question 1

#### define and create AOI - Palo, Iowa

```{r Q1, message=FALSE, warning=FALSE}

# read uscities.csv data into R, make it a sf object, and filter to Goleta
paloIowa <- read_csv("~/github/geog-176A-labs/data/uscities.csv") %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%
  filter(city == "Palo")%>%
  # transform to CRS:5070 and create a 5 km buffer
  st_transform(5070) %>%
  st_buffer(5000) %>%
  # get bounding box of buffer and make bbox object a sfc and then sf object
  st_bbox() %>%
  st_as_sfc() %>%
  st_as_sf()

```
***
```{r no, include=FALSE}

newbox <- paloIowa %>%
  st_transform(4326) %>%
  st_bbox()

landstatdata <- getlandsat::lsat_scenes()

filtered <- landstatdata %>%
  filter(as.Date(acquisitionDate) == "2016-09-26") %>%
  filter(min_lat <= newbox$ymin) %>%
  filter(min_lon <= newbox$xmin) %>%
  filter(max_lon >= newbox$xmax) %>%
  filter(max_lat >= newbox$ymax)

```

### Question 2

#### download/cache files and stack raster files

```{r q1 part 2, warning=FALSE, message=FALSE}

data <- write.csv(filtered, file = "C:/Users/Catherine/Documents/github/geog-176A-labs/data/palofloodscene.csv")

metadata <- read_csv("~/github/geog-176A-labs/data/palofloodscene.csv")

pfiles <- lsat_scene_files(metadata$download_url) %>%
  filter(grepl(paste0('B',1:6,'.TIF$', collapse = '|'), file)) %>%
  arrange(file) %>%
  pull(file)

############################

st <- sapply(pfiles, lsat_image)

b <- stack(st) %>%
  setNames(paste0('band', 1:6))

```

***
- What are the dimensions of your stacked image? What is the CRS?

7,811 rows with 7,861 columns and 6 layers for a total of 59,996,291 cells.

- What is the cell resolution?

WGS84 

- What is the cell resolution?

30m by 30m


***


#### crop raster stack to AOI

```{r q2, warning=FALSE, message=FALSE}


AOIcrop <- paloIowa %>%
  st_as_sf() %>%
  st_transform(crs(b))

cropped <- crop(b, AOIcrop)

```
***

- What are the dimensions of your cropped image stack?

340 rows with 346 columns and 6 layers for a total of 117,640 cells.

- What is the CRS?

WGS84

- What is the cell resolution?

30m by 30m

***

### Question 3

#### Rename raster stack to correspaoning band names and plot band combinations

```{r q3, warning=FALSE, message=FALSE}

rasterstack <- cropped %>%
  setNames(c('Coastal aerosol', 'Blue', 'Green', 'Red', 'Near Infrared', 'SWIR 1'))

op <- par(mfrow = c(1, 2),
          pty = "s")

plotRGB(rasterstack, 4, 3, 2)
plotRGB(rasterstack, 5, 4, 2)
plotRGB(rasterstack, 5, 6, 4)
plotRGB(rasterstack, 1, 2, 6)


##############################


plotRGB(rasterstack, 4, 3, 2, stretch = 'hist')
plotRGB(rasterstack, 5, 4, 2, stretch = 'lin')
plotRGB(rasterstack, 5, 6, 4, stretch = 'hist')
plotRGB(rasterstack, 1, 2, 6, stretch = 'lin')

par(op)


```

***
Describe the purpose of applying a color stretch.

A color stretch increases the color range of the pixels and improves the contrast and clarity.

***

### Question 4

#### Create 5 new rasters using formulas for NDVI, NDWI, MNDWI, WRI and SWI

```{r q4, warning=FALSE, message=FALSE}

normalized_dvi <- (rasterstack$Near.Infrared - rasterstack$Red) / (rasterstack$Near.Infrared + rasterstack$Red)

normalized_dwi <-  (rasterstack$Green - rasterstack$Near.Infrared) / (rasterstack$Green + rasterstack$Near.Infrared)

modified_ndwi <-  (rasterstack$Green - rasterstack$SWIR.1) / (rasterstack$Green + rasterstack$SWIR.1)

waterratioindex <- (rasterstack$Green + rasterstack$Red) / (rasterstack$Near.Infrared + rasterstack$SWIR.1)

simplewaterindex <- 1 / ( sqrt(rasterstack$Blue - rasterstack$SWIR.1) )


stackedraster <- stack(normalized_dvi, normalized_dwi, modified_ndwi, waterratioindex, simplewaterindex) %>%
  setNames(c('Normalized Difference Vegetation',
             'Normalized Difference Water',
             'Mod Normalized Difference Water',
             'Water Ratio', 'Simple Water'))

plot(stackedraster, col = colorRampPalette(c("blue", "white", "red"))(256))


```

***

# Describe the 5 images. How are they simular and where do they deviate?

***

#### create flooding thresholds for the rasters

```{r q4 part2, message=FALSE, warning=FALSE}

threshold_ndvi <- function(x) {
  ifelse( x <= 0, 1, 0)
}

threshold_ndwi <- function(x) {
  ifelse( x >= 0, 1, 0)
}

threshold_mndwi <- function(x) {
  ifelse( x >= 0, 1, 0)
}

threshold_wri <- function(x) {
  ifelse( x >= 1, 1, 0)
}

threshold_swi <- function(x) {
  ifelse( x <= 5, 1, 0)
}


binary_ndvi <- calc(normalized_dvi, threshold_ndvi)

binary_ndwi <- calc(normalized_dwi, threshold_ndwi)

binary_mndwi <- calc(modified_ndwi, threshold_mndwi)

binary_wri <- calc(waterratioindex, threshold_wri)

binary_swi <- calc(simplewaterindex, threshold_swi)



binarystack <- stack(c(binary_ndvi, binary_ndwi, binary_mndwi, binary_wri, binary_swi)) %>%
  setNames(c('Normalized Difference Vegetation',
             'Normalized Difference Water',
             'Mod Normalized Difference Water',
             'Water Ratio', 'Simple Water'))

plot(binarystack, col = colorRampPalette(c("white","blue"))(256))


binary_ndvi[is.na(values(binary_ndvi))] <- 0
binary_ndwi[is.na(values(binary_ndwi))] <- 0
binary_mndwi[is.na(values(binary_mndwi))] <- 0
binary_wri[is.na(values(binary_wri))] <- 0
binary_swi[is.na(values(binary_swi))] <- 0

```

***

### Question 5


#### Extract values from 6-band raster stack  
```{r q5, message=FALSE, warning=FALSE}

set.seed(20200907)


values <- getValues(rasterstack)

dim(values)
dim(rasterstack)

idx <- which(!is.na(values))
values <- na.omit(values)


```

***

What do the dimensions of the extracted values tell you about how the data was extracted?

Before the values were extracted the raster had 340 rows and 346 columns and 6 layers. After the values were extracted it had 117,640 rows and 6 columns. The layers became columns and the number of cells became the number of rows. 

***

#### create new kmeans raster object

```{r q4 part 2, message=FALSE, warning=FALSE}

kmeans <- kmeans(values, centers = 10)

kmeanraster <- rasterstack$Coastal.aerosol
values(kmeanraster) <- NA
kmeanraster[idx] <- kmeans$cluster


##############################


table <- table(getValues(binary_ndvi), getValues(kmeanraster))


kmeanraster[kmeanraster != which.max(table)] <- 0
kmeanraster[kmeanraster != 0] <- 1


binarystack <- raster::addLayer(binarystack, kmeanraster)


plot(binarystack, col = colorRampPalette(c("white", "blue"))(256))


```

***
### Question 6


#### calculate the total area of floods and visualize uncertainty in the classifications
```{r q6, message=FALSE, warning=FALSE}

area <- cellStats(binarystack, 'sum') * (30^2)

kable_styling(kable(data.frame(area), col.names = ('Area'),
                    caption = 'Total Area of the Flooded Cells'))


####################

summedfloods <- calc(binarystack, fun = sum) 
plot(summedfloods, col = blues9, title = "Flood Uncertanity")

```

***

#### create pannable map

```{r q6 part2, message=FALSE, warning=FALSE}

summedfloods[summedfloods == 0] <- NA


pal <- mapviewPalette("mapviewTopoColors")
mapview(summedfloods, col = pal)

```

