---
title: "Geography 176A"
subtitle: "Lab 03: COVID-19 Pandemic"
author: "[Catherine Rauch](https://catherinerauch.github.io)"
date: "8/13/2020"
output: html_document
html_document:
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r packages, include=FALSE}
library(tidyverse) # data wrangling and visualization
library(knitr) # make nice tables
library(readxl) # read excel files
library(zoo) # rolling averages
```

```{r Q1, warning=FALSE,echo=FALSE, message=FALSE}

coviddata <- read_csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv')

stateofinterest <- 'California'

```

```{r q1 cont, warning=FALSE, message=FALSE}


# Question 1

addednewcases <- coviddata %>% # filter data by state of interest
  filter(state == stateofinterest) %>%
  group_by(county, date) %>%
  summarise(cases = sum(cases), fips) %>%
  mutate(newCases = cases - lag(cases)) %>% # creates new column with daily new cases
  ungroup()

mostnewcases <- addednewcases %>% # finds top 5 counties with most new cases per day
    filter(date == max(date)) %>%
    slice_max(newCases, n = 5) %>%
    select(county, newCases)

mostcumulativecases <- addednewcases %>% # finds top 5 counties with most cumulative cases
  filter(date == max(date)) %>%
  slice_max(cases, n = 5) %>%
  select(county, cases)

knitr::kable(mostnewcases, caption = "Most New Cases California Counties", col.names = c("County", "New Cases"),
             format.args = list(big.mark = ","))


```

**Los Angeles, San Bernardino, Riverside, Fresno and Orange County** are the counties with the most new cases reported daily. Both Los Angeles and San Bernado reported over a thousand.

```{r table2}


knitr::kable(mostcumulativecases, caption = "Most Cumulative Cases California Counties",
             col.names = c("County", "Total Cases"), format.args = list(big.mark = ","))


```

**Los Angeles, Riverside, Orange, San Bernardino, and San Diego** are the counties with the most total reported cases, this could be due to be being the largest 5 counties in California. Los Angeles is the only reporting more than 50,000 cases currently. 


```{r Q1 part2, warning=FALSE, message=FALSE}


PopulationEstimates <- read_excel("~/github/geog-176A-labs/data/PopulationEstimates.xls", skip = 2)
# dim(PopulationEstimates) # examine dimensions of excel file
# head(PopulationEstimates,5)

populationestimates <- PopulationEstimates %>% # select only necessary columns
  select(fips = "FIPStxt", state = "State", Area_Name, pop2019 = "POP_ESTIMATE_2019")


CovidandPopulation <- addednewcases %>%
  inner_join(populationestimates, by = "fips" ) # join to previous data


mostnewcasespc <- CovidandPopulation %>% # finds top 5 counties with daily new cases per capita
  filter(date == max(date)) %>%
  group_by(county) %>%
  mutate(casepercap = cases / pop2019) %>%
  ungroup() %>%
  slice_max(casepercap, n = 5) %>%
  select(county, casepercap)


mostcumulativecasespc <-CovidandPopulation %>% # finds top 5 counties with cumulative cases per capita
  filter(date == max(date)) %>%
  group_by(county) %>%
  mutate(newcasepercap = newCases / pop2019) %>%
  ungroup() %>%
  slice_max(newcasepercap, n = 5) %>%
  select(county, newcasepercap)

knitr::kable(mostnewcasespc, caption = "Most New Cases California Counties Per Capita", col.names = c("County", "New Cases Per Capita"))


```

Examining the same data with respect to population size, it can be seen that the 5 highest counties of new daily cases per capita are **Imperial, Kings, Kern, Tulare and Merced County.** While most total cases per capita are **Inyo, Stanislaus, Fresno, Kings, and San Bernardino County.**

```{r Q1 part4, warning=FALSE, message=FALSE}


knitr::kable(mostcumulativecasespc, caption = "Most Cumulative Cases California Counties Per Capita",
             col.names = c("County", "Total Cases Per Capita"))

numofdays <- 14

greaterthan100 <- CovidandPopulation %>%
  filter(date > (max(date) - numofdays)) %>%
  group_by(county, date) %>%
  summarise(ncases =  100000 * (sum(newCases) / pop2019)) %>%
  filter(ncases > 100)

knitr::kable(greaterthan100, caption = "California Counties with > 100 cases within 2 weeks ", col.names = c("County", "Date", "Cases Per 100,000 Residents"))


```

##### 15 counties have more than 100 cases per 100,000 residents in the past 14 days, that leaves **43** counties deemed as safe.

```{r Q2, warning=FALSE, message=FALSE}


## Question 2

statesofinterest <- c("New York", "California", "Louisiana","Florida")

coviddatastate <- coviddata %>%
  filter(state %in% statesofinterest) %>%
  group_by(state, date) %>%
  summarise(cases = sum(cases)) %>%
  mutate(newCases = cases - lag(cases),  roll7 = rollmean(newCases, 7, fill = NA, align="right")) %>% # creates new column withe daily new cases
  ungroup()

  coviddatastate %>%
  ggplot(aes(x = date, y = newCases)) +  geom_col(aes(y = newCases, col = state))+
  geom_line(aes(col = state)) + geom_line(aes(y = roll7), size = 1) +
  facet_grid(~state, scale = "free_y", space = "free_y") + ggthemes::theme_solarized()+
  theme(legend.position = "bottom") +
  theme(legend.position = "NA") +
  labs(title = "Daily new cases by State", y = "Daily New Count", x = "Date",
       caption = "Lab 02")
  
```

```{r Q2 part 2,warning=FALSE, message=FALSE}

coviddatastate %>%
  inner_join(populationestimates, by = c("state" = "Area_Name") ) %>%
  group_by(state, date) %>%
  summarise(newcases = sum(newCases), pop2019, .groups = "drop") %>%
  mutate(newcasepercap = (newcases / pop2019), roll7 = rollmean(newcasepercap, 7, fill = NA, align="right")) %>%
  ggplot(aes(x = date, y = newcasepercap)) +
  geom_col(aes(y = newcasepercap, col = state))+
  geom_line(aes(col = state)) + geom_line(aes(y = roll7), size = 1)+
  facet_grid(~state) + ggthemes::theme_solarized() + ggthemes::scale_color_gdocs() +
  theme(legend.position = "none") +
  labs(title = "Daily new cases Per Capita by State", y = "Daily New Count Per Capita", x = "Date",
       caption = "Lab 02")


```


##### This graph was scaled by population, and it can be seen that Louisiana's situation is more dire while California's appear better. 
##### California has the largest population among these states and after scaling, it can be seen that they have had the smallest reported spike in daily cases. 
##### New York and Florida are close together in population and appear to have a similar sized spike in daily cases. 
##### Whereas before the scaling it appeared Florida had the largest spike, and California and New York's was similar.
##### Although New York's was in April while Florida's happened in July.
##### Louisiana has the smallest population by far with roughly 4.6 million (New York, California and Florida have a population greater than 19 million) but, after scaling, has the largest spike in daily new cases per capita.

